---
title: "Abalone Ring Predictor"
author: "Cristina Castaneda"
date: "5/6/2022"
output:
  github_document: default
  pdf_document: default
---

```{r warning=FALSE, message=FALSE}
# load packages
library(tidyverse)
library(rpart)
library(randomForest)
library(vip)
library(ggplot2)
library(AppliedPredictiveModeling)
```




```{r}
data("abalone")
```

## EDA


* There is a positive correlation between Rings and the variables WholeWeight, Diameter, ShuckedWeight, and ShellWeight. 
* The average diameter for the abalone is 0.41. 
* The average ShuckedWeight is 0.36. 
* The average WholeWeight is 0.83. 
* When comparing Rings by type it appears that females have the largest amount of Rings. 
* There is an outlier in the female data, there is a female abalone that has almost 30 rings. 
* It looks like the most common number of Rings for females and males is between 8-15. 
* For infants the most common number of Rings appears to be between 4-10. 

```{r}
head(abalone)
```


```{r}
dim(abalone)
```

```{r}
summary(abalone)
```

```{r}
ggplot(abalone, aes(WholeWeight, Rings)) + geom_point()
```

```{r}
ggplot(abalone, aes(Diameter,Rings)) + geom_point()
```

```{r}
ggplot(data = abalone, aes(ShuckedWeight, Rings)) + geom_point()
```

```{r}
ggplot(data = abalone, aes(ShellWeight, Rings, color = Type)) + geom_point(size = 0.3, alpha = 0.7) + geom_smooth(method = "lm", se= FALSE)+ facet_wrap(~Type) 
```




```{r}
# randomly split data into a 70% training and 30% test set
set.seed(500)
n <- nrow(abalone)
train_index <- sample(1:n, round(0.7*n)) 
abalone_train <- abalone[train_index, ]
abalone_test <- abalone[-train_index, ]
```




```{r}
# summary linear model abolone ring prediction
lm = lm(Rings ~ ., data = abalone_train)
summary(lm)
```



```{r}
# fit tree model on training set
t1 = rpart(Rings ~ . , abalone_train, method = "anova")
```


```{r}
# plot tree model 
par(cex = 0.7, xpd = NA)
plot(t1, uniform = TRUE)
text(t1, use.n = TRUE, pretty = 0)
```


The two most important variables are ShuckedWeight and Type. 

```{r}
# random forest for predicting rings on abalone training data 
rf1 = randomForest(Rings ~ ., abalone_train, importance = TRUE)
```



```{r}
# variable importance plot generated from random forest
vip(rf1, num_features = 9, geom = "point", include_type = TRUE)
```




 
```{r}
RMSE = function(y, y_hat) 
  {
  sqrt(mean((y-y_hat)^2))
  }

```


```{r}
# make predictions on the test set for each model

pred_rf1 = predict(rf1, newdata = abalone_test)
pred_lm = predict(lm, newdata = abalone_test)
pred_t1 = predict(t1, newdata = abalone_test)
```


```{r}
# RMSE for random forest
RMSE(abalone_test$Rings, pred_rf1)
```

```{r}
#RMSE for MLR model
RMSE(abalone_test$Rings, pred_lm)
```

```{r}
# RMSE for tree model
RMSE(abalone_test$Rings, pred_t1)
```
```{r}
# R^2 random forest
cor(abalone_test$Rings, pred_rf1)^2
```
```{r}
# R^2 linear model
cor(abalone_test$Rings, pred_lm)^2
```

```{r}
# R^2 tree model
cor(abalone_test$Rings, pred_t1)^2
```



```{r}
## predict dataframe and graph of plot vs predicted

pred_df = function(model)
{
  data.frame(
  Actual = abalone_test$Rings,
  Predicted = model
  )
}

actualvspred = function(df){
  ggplot(df, aes(x=Actual, y = Predicted))+
    geom_point(size = 0.5, alpha = 0.5)+
    geom_abline(intercept = 0, slope = 1, color = "red")
}

```


```{r}
## plot of predicted vs actual for random forest 
df1 = pred_df(pred_rf1)
actualvspred(df1)
```


```{r}
## plot of predicted vs actual for tree model
df2 = pred_df(pred_t1)
actualvspred(df2)
```


```{r}
## plot of predicted vs actual for linear regression model
df3 = pred_df(pred_lm)
actualvspred(df3)
```


